---
title: "Getting started with rze"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rze}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

rze finds where net diversification rate shifts on a phylogeny of extant taxa, and reports the rate for each part of the tree it identifies. This vignette covers the common case: you have a tree, you want the shifts and rates, and you want to use the calibration table that ships with the package so you can skip the calibration step entirely.

## The one-line version

```{r}
library(rze)

set.seed(1)
sim <- simulate_multi_shift_tree(
  n_background = 150,
  shifts = list(
    list(n_shifted = 30, lambda_ratio = 8),
    list(n_shifted = 20, lambda_ratio = 4)
  ),
  epsilon = 0.2, rho = 1
)
my_tree <- sim$tree

result <- rze(my_tree, rho = 0.8)
result
```

`rho` is your sampling fraction, the proportion of the group's species that made it into the tree. If you sampled 80 of an estimated 100 species, `rho = 0.8`. Set it as accurately as you can, since detection depends on it.

That single call runs the whole thing. It picks up the shipped calibration table automatically, searches the tree for clade-localized shifts, and estimates a net diversification rate for each regime. You do not run a calibration step, the table is already there.

## What you get back

Printing the result gives you the detected shifts and a rate for each regime:

```{r}
result
#> rze diversification shift analysis
#>   tree: 220 tips
#>   sampling fraction (rho): 0.80
#>   detected shifts: 2
#>
#>      regime node n_tips net_diversification boundary_mu
#>  background   NA    170              0.0912       FALSE
#>    shift_1  245     30              0.4821       FALSE
#>    shift_2  268     20              0.2765       FALSE
```

The background row is the rest of the tree, the part with no shift. Each `shift_` row is a clade the search flagged as diversifying differently, with its own rate. `net_diversification` is speciation minus extinction, which is the rate you can actually trust from an extant-only tree. `boundary_mu` flags a regime whose extinction estimate ran to zero, when that happens the net rate is still fine but you should not read anything into the split between speciation and extinction there.

## Seeing it

The main plot colors every branch by its regime's rate, so the fast and slow parts of the tree are obvious at a glance, and marks each detected shift:

```{r}
plot(result)
```

The rate comparison plot lines up each detected clade against the background rate, which is usually the figure that answers the question you actually care about, which clades stand out:

```{r}
plot_regime_rates(result)
```


## Whale dataset

```{r eval=FALSE, include=TRUE}
data(whales, events.whales)
my_tree <- force.ultrametric(whales)
result <- rze(my_tree, rho = 0.8)
result
plot(result)
plot_regime_rates(result)
```


## A few things worth knowing

Detection power depends on tree size. Small trees, roughly under 30 tips, have low power even when a shift is real and large. Recovery gets reliable as trees grow into the hundreds of tips. If rze finds nothing on a small tree, that is often a power limit, not evidence that no shift exists.

rze reports net diversification, not speciation and extinction separately. That is deliberate. On a tree of living species you cannot cleanly separate the two, so reporting them as if you could would be dishonest. Net diversification is the quantity the data supports.

The sampling fraction here is a single global value. If your sampling is very uneven across the tree, see the calibration vignette for how to handle that more carefully.
